---
output:
  github_document:
    html_preview: false
---

```{r setup1, include=FALSE}
set.seed(1)
```

[![CRAN](https://www.r-pkg.org/badges/version/anndata)](https://cran.r-project.org/package=anndata)
[![Build status](https://github.com/rcannood/anndata/workflows/R-CMD-check/badge.svg)](https://github.com/rcannood/anndata/actions)

# anndata - Annotated Data

[`anndata`](https://anndata.readthedocs.io/en/latest/) is a commonly used Python package 
for keeping track of data and learned annotations, and can be used to read from and write to the h5ad 
file format. It is also the main data format used in the scanpy python package [@wolf_scanpylargescalesinglecell_2018].

![anndata](man/readme_files/anndata_for_r.svg)

However, using scanpy/anndata in R can be a major hassle. When trying to read an h5ad file,
R users could approach this problem in one of two ways. A) You could read in the file manually
(since it's an H5 file), but this involves a lot of manual work and a lot of understanding
on how the h5ad and H5 file formats work (also, expect major headaches from cryptic hdf5r bugs).
Or B) interact with scanpy and anndata through reticulate, but run into issues converting
some of the python objects into R.

We recently published [`anndata`](https://cran.r-project.org/package=anndata) on CRAN,
which is an R wrapper for the Python package -- with some syntax sprinkled on top to make
R users feel more at home. 

The R anndata is still under active development at [github.com/rcannood/anndata](https://github.com/rcannood/anndata). 
If you encounter any issues, feel free to post an issue on GitHub!

## Installation

Installing `anndata` is not particularly hard. You do need R and Python to be installed.
If you don't have a working installation of Python, you can let reticulate install Miniconda.

```r
# install the R anndata package
install.packages("anndata")

# skip this if you wish to use the local python installation
reticulate::install_miniconda()

# skip this if anndata is already installed
anndata::install_anndata()
```

## Getting started

The API of `anndata` is very similar to its Python counterpart. 
Check out `?anndata` for a full list of the functions provided by this package.

`AnnData` stores a data matrix `X` together with annotations
of observations `obs` (`obsm`, `obsp`), variables `var` (`varm`, `varp`),
and unstructured annotations `uns`. 

Here is an example of how to create an AnnData object with 2 observations and 3 variables.

```{r}
library(anndata)

ad <- AnnData(
  X = matrix(1:6, nrow = 2),
  obs = data.frame(group = c("a", "b"), row.names = c("s1", "s2")),
  var = data.frame(type = c(1L, 2L, 3L), row.names = c("var1", "var2", "var3")),
  layers = list(
    spliced = matrix(4:9, nrow = 2),
    unspliced = matrix(8:13, nrow = 2)
  ),
  obsm = list(
    ones = matrix(rep(1L, 10), nrow = 2),
    rand = matrix(rnorm(6), nrow = 2),
    zeros = matrix(rep(0L, 10), nrow = 2)
  ),
  varm = list(
    ones = matrix(rep(1L, 12), nrow = 3),
    rand = matrix(rnorm(6), nrow = 3),
    zeros = matrix(rep(0L, 12), nrow = 3)
  ),
  uns = list(
    a = 1, 
    b = data.frame(i = 1:3, j = 4:6, value = runif(3)),
    c = list(c.a = 3, c.b = 4)
  )
)

ad
```

You can read the information back out using the `$` notation.
```{r}
ad$X
ad$obs
ad$obsm[["ones"]]
ad$layers[["spliced"]]
ad$uns[["b"]]
```

### Using scanpy 

A nice side-effect of loading this package is that it is now super easy to 
interact with scanpy through these AnnData objects.

```{r}
library(reticulate)

ad$X

sc <- import("scanpy")
sc$pp$normalize_per_cell(ad)

ad$X
```

### Reading / writing AnnData objects

Read from h5ad format:
```{r}
read_h5ad("example_formats/pbmc_1k_protein_v3_processed.h5ad")
```

### Creating a view
You can use any of the regular R indexing methods to subset the `AnnData` object.
This will result in a 'View' of the underlying data without needing to store 
the same data twice.
```{r}
view <- ad[, 2]
view
view$is_view

ad[,c("var1", "var2")]
ad[-1, ]
```

### AnnData as a matrix
The `X` attribute can be used as an R matrix:
```{r}
ad$X[,c("var1", "var2")]
ad$X[-1, , drop = FALSE]
ad$X[, 2] <- 10
```

You can access a different layer matrix as follows:
```{r}
ad$layers["unspliced"]
ad$layers["unspliced"][,c("var2", "var3")]
```

### Note on state
If you assign an AnnData object to another variable and
modify either, both will be modified:
```{r}
ad2 <- ad

ad$X[,2] <- 10

list(ad = ad$X, ad2 = ad2$X)
```

This is standard Python behaviour but not R. In order to 
have two separate copies of an AnnData object, use the `$copy()` function:
```{r}
ad3 <- ad$copy()

ad$X[,2] <- c(3, 4)

list(ad = ad$X, ad3 = ad3$X)
```

## Future work

In some cases, this package may still act more like a Python package rather than an R package.
Some more helper functions and helper classes need to be defined in order to fully encapsulate
`AnnData()` objects. Examples are:

```{r}
ad$chunked_X(1)
```

Following functionality has not been tested:
```r
ad$rename_categories(...)
ad$strings_to_categoricals(...)
```

Currently not implemented are the `read_zarr()` and `ad$write_zarr()` functions. 
I need some example data to test this functionality.

## References

## Latest changes
Check out `news(package = "anndata")` or [NEWS.md](NEWS.md) for a full list of changes.

<!-- This section gets automatically generated from NEWS.md -->

```{r news, echo=FALSE, results="asis"}
cat(dynutils::recent_news())
```

