---
output:
  github_document:
    html_preview: false
---

```{r setup1, include=FALSE}
set.seed(1)
```

[![Stars](https://img.shields.io/github/stars/theislab/anndata?logo=GitHub&color=yellow)](https://github.com/theislab/anndata/stargazers)
[![CRAN](https://www.r-pkg.org/badges/version/anndata)](https://cran.r-project.org/package=anndata)
[![Build status](https://github.com/rcannood/anndata/workflows/R-CMD-check/badge.svg)](https://github.com/rcannood/anndata/actions)

# anndata - Annotated Data

`anndata` provides a scalable way of keeping track of data
and learned annotations, and can be used to read from and write to the h5ad 
file format.

![anndata](http://falexwolf.de/img/scanpy/anndata.svg)

This package is, in essense, an R wrapper for the similarly named Python package
[`anndata`](https://anndata.readthedocs.io/en/latest/), with some added functionality
to support more R-like syntax. The version number of the anndata R package is synced
with the version number of the python version.


## Installation

```r
# install the R anndata package
install.packages("anndata")

# run this only if you do not already have an installation of miniconda
reticulate::install_miniconda()

# install the Python anndata package
anndata::install_anndata()
```

## Getting started

The API of `anndata` is very similar to its Python counterpart. 
Check out `?anndata` for a full list of the functions provided by this package.

`AnnData` stores a data matrix `X` together with annotations
of observations `obs` (`obsm`, `obsp`), variables `var` (`varm`, `varp`),
and unstructured annotations `uns`. 

Here is an example of an AnnData object with 2 observations and 3 variables.

```{r}
library(anndata)

ad <- AnnData(
  X = matrix(1:6, nrow = 2),
  obs = data.frame(group = c("a", "b"), row.names = c("s1", "s2")),
  var = data.frame(type = c(1L, 2L, 3L), row.names = c("var1", "var2", "var3")),
  layers = list(
    spliced = matrix(4:9, nrow = 2),
    unspliced = matrix(8:13, nrow = 2)
  ),
  obsm = list(
    ones = matrix(rep(1L, 10), nrow = 2),
    rand = matrix(rnorm(6), nrow = 2),
    zeros = matrix(rep(0L, 10), nrow = 2)
  ),
  varm = list(
    ones = matrix(rep(1L, 12), nrow = 3),
    rand = matrix(rnorm(6), nrow = 3),
    zeros = matrix(rep(0L, 12), nrow = 3)
  ),
  uns = list(
    a = 1, 
    b = data.frame(i = 1:3, j = 4:6, value = runif(3)),
    c = list(c.a = 3, c.b = 4)
  )
)

ad
```

You can read the information back out using the `$` notation.
```{r}
ad$X

ad$obs
ad$var

ad$obsm["ones"]
ad$varm["rand"]

ad$layers["unspliced"]
ad$layers["spliced"]

ad$uns["b"]
```

### Reading / writing AnnData objects

Read from h5ad format:
```{r}
read_h5ad("example_formats/pbmc_1k_protein_v3_processed.h5ad")
```


### AnnData as a matrix
An `AnnData` object can be used as an R matrix:
```{r}
ad[,c("var1", "var2")]
ad[-1, , drop = FALSE]
ad[, 2] <- 10
```

You can simply use `ad[]` to get quick access to the `X` matrix, or add in 
`layer="unspliced"` to switch to a different layer.
```{r}
ad[]
ad[layer="unspliced"]
ad[,c("var2", "var3"),layer="unspliced"]
```

### Note on state
If you assign an AnnData object to another variable and
modify either, both will be modified:
```{r}
ad2 <- ad

ad$X[,2] <- 10

list(ad = ad$X, ad2 = ad2$X)
```

This is standard Python behaviour but not R. In order to 
have two separate copies of an AnnData object, use the `$copy()` function:
```{r}
ad3 <- ad$copy()

ad$X[,2] <- c(3, 4)

list(ad = ad$X, ad3 = ad3$X)
```

## Future work

In some cases, this package may still act more like a Python package rather than an R package.
Some more helper functions and helper classes need to be defined in order to fully encapsulate
`AnnData()` objects. Examples are:

```{r}
ad$layers
ad$chunked_X(1)
```

Following functionality has not been tested:
```r
ad$rename_categories(...)
ad$strings_to_categoricals(...)
```

Currently not implemented are the `read_zarr()` and `ad$write_zarr()` functions.

## Latest changes
Check out `news(package = "anndata")` or [NEWS.md](NEWS.md) for a full list of changes.

<!-- This section gets automatically generated from NEWS.md -->

```{r news, echo=FALSE, results="asis"}
cat(dynutils::recent_news())
```

