---
output:
  github_document:
    html_preview: false
bibliography: man/readme_files/references.bib
---

```{r setup1, include=FALSE}
library(tidyverse)
set.seed(1)
```

# anndata for R

<!-- badges: start -->
[![CRAN](https://www.r-pkg.org/badges/version/anndata)](https://cran.r-project.org/package=anndata)
[![CRAN Downloads](https://cranlogs.r-pkg.org/badges/anndata)](https://cran.r-project.org/package=anndata)
[![R-CMD-check](https://github.com/dynverse/anndata/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/dynverse/anndata/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://app.codecov.io/gh/dynverse/anndata/branch/main/graph/badge.svg)](https://app.codecov.io/gh/dynverse/anndata?branch=main)
[![Lifecycle: superseded](https://img.shields.io/badge/lifecycle-superseded-blue.svg)](https://lifecycle.r-lib.org/articles/stages.html#superseded)
<!-- badges: end -->

> **⚠️ This package is superseded by [anndataR](https://anndataR.scverse.org).**
>
> `anndataR` provides a pure R implementation of the `AnnData` data structure — no Python
> required. It reads and writes `.h5ad` files natively and supports conversion to/from
> `SingleCellExperiment` and `Seurat` objects. New users should install `anndataR` from
> Bioconductor instead:
>
> ```r
> BiocManager::install("anndataR")
> ```
>
> Existing users of `anndata` can follow the
> [migration guide](https://anndata.dynverse.org/articles/migration_to_anndataR.html)
> for guidance on switching to `anndataR`.

`anndata` for R is a `reticulate` wrapper for the Python `anndata` package.

## Installation

> **Note:** New projects should use [anndataR](https://anndataR.scverse.org) instead.
> See the [migration guide](https://anndata.dynverse.org/articles/migration_to_anndataR.html).

You can install `anndata` for R from CRAN as follows:

```r
install.packages("anndata")
```

Normally, reticulate should take care of installing Miniconda and the Python anndata.

If not, try running:

```r
reticulate::install_miniconda()
anndata::install_anndata()
```

## Getting started

The API of anndata for R is very similar to its Python counterpart. Here is an example:

```{r}
library(anndata)

ad <- read_h5ad("example_formats/pbmc_1k_protein_v3_processed.h5ad")

ad

Matrix::rowMeans(ad$X[1:10, ])
```


See `?anndata` for a full list of the functions provided by this package.
Check out any of the other vignettes by clicking any of the links below:

```{r vignettes, results='asis', echo=FALSE}
walk(
  list.files("vignettes", pattern = "*.Rmd", recursive = TRUE),
  function(file) {
    title <-
      read_lines(paste0("vignettes/", file)) %>%
      keep(~ grepl("^title: ", .)) %>%
      gsub("title: \"(.*)\"", "\\1", .)
    vignette_name <- gsub("\\.Rmd", "", file)
    html_name <- gsub("\\.Rmd", ".html", file)

    cat("* [", title, "](https://anndata.dynverse.org/articles/", html_name, ")\n", sep = "")
  }
)
```

## Future work

This package is no longer under active development as it has been superseded by
[anndataR](https://anndataR.scverse.org). Bug fixes may still be
applied, but no new features are planned.

## References
