---
title: "Migrating from anndata to anndataR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Migrating from anndata to anndataR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r check_on_cran, message=FALSE, warning=FALSE, echo=FALSE}
on_cran <- !identical(Sys.getenv("NOT_CRAN"), "true")

if (on_cran) {
  knitr::opts_chunk$set(eval = FALSE)
  knitr::asis_output(paste0(
    "<span style=\"color: red;\">**Note:** The outputs of this vignette are not rendered on CRAN due to package size limitations. ",
    "Please check the [Migration to anndataR](https://anndata.dynverse.org/articles/migration_to_anndataR.html) ",
    "vignette in the package documentation.</span>"
  ))
}
```

## Why migrate?

The `anndata` R package (this package) is superseded by
[anndataR](https://anndataR.scverse.org), available on Bioconductor.

| Feature | `anndata` (CRAN) | `anndataR` (Bioconductor) |
|---|---|---|
| **Python dependency** | Required | Optional (only needed for `ReticulateAnnData`) |
| **h5ad I/O** | Via Python | Native R via `rhdf5` (preferred), or via Python reticulate |
| **In-memory backend** | Python-backed | Native R (`InMemoryAnnData`) |
| **HDF5-backed backend** | Python-backed | Native R (`HDF5AnnData`) |
| **Reticulate-backed backend** | Always | Optional (`ReticulateAnnData`) |
| **Seurat interop** | Not supported | `as_Seurat()` / `as_AnnData()` |
| **SingleCellExperiment interop** | Not supported | `as_SingleCellExperiment()` / `as_AnnData()` |
| **Distribution** | CRAN | Bioconductor |

The preferred backends (`InMemoryAnnData`, `HDF5AnnData`) require no Python.
For users who already have Python `anndata` installed, the `ReticulateAnnData`
backend offers a low-friction starting point; see
[below](#using-reticulateanndata-as-a-stepping-stone).

## Installation

```{r install}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}
BiocManager::install("anndataR")

# Required for native h5ad reading/writing:
BiocManager::install("rhdf5")

# Optional: SingleCellExperiment conversion
BiocManager::install("SingleCellExperiment")

# Optional: Seurat conversion
install.packages("SeuratObject")
```

## Side-by-side migration guide

### Creating an AnnData object

The `AnnData()` constructor is identical in both packages. The only difference
is that `anndataR` returns an `InMemoryAnnData` (a pure-R object) whereas
`anndata` returned an `AnnDataR6` wrapping a Python object.

```{r create}
library(anndataR)

ad <- AnnData(
  X = matrix(1:6, nrow = 2),
  obs = data.frame(group = c("a", "b"), row.names = c("s1", "s2")),
  var = data.frame(type = c(1L, 2L, 3L), row.names = c("var1", "var2", "var3")),
  layers = list(spliced = matrix(4:9, nrow = 2)),
  uns = list(a = 1)
)
```

### Reading and writing h5ad files

The function signatures are identical. `anndataR` reads natively without Python.

```{r read-write}
# Reading
ad <- read_h5ad("path/to/file.h5ad")                               # InMemoryAnnData (default)
ad <- read_h5ad("path/to/file.h5ad", as = "HDF5AnnData")           # disk-backed, low memory
sce <- read_h5ad("path/to/file.h5ad", as = "SingleCellExperiment")
obj <- read_h5ad("path/to/file.h5ad", as = "Seurat")

# Writing (works identically to anndata)
write_h5ad(ad, "path/to/output.h5ad")
```

### Slot access and subsetting

Both packages use the same `$` notation and bracket subsetting syntax.

```{r slots}
ad$X; ad$obs; ad$var; ad$obsm; ad$varm; ad$obsp; ad$varp; ad$layers; ad$uns

# Subsetting returns an AnnDataView (native R) rather than a Python-backed view
subset <- ad[1:5, ]
subset <- ad[, c("var1", "var2")]
subset <- ad[ad$obs$group == "a", ]
concrete <- subset$as_InMemoryAnnData()  # materialise to a concrete object
```

### Interoperability (new in anndataR)

```{r interop}
# AnnData <-> SingleCellExperiment
sce <- ad$as_SingleCellExperiment()
ad  <- as_AnnData(sce)

# AnnData <-> Seurat
obj <- ad$as_Seurat()
ad  <- as_AnnData(obj)
```

### Using ReticulateAnnData as a stepping stone

`anndataR`'s `ReticulateAnnData` backend wraps a Python `anndata.AnnData` object
via reticulate, implementing the same `AbstractAnnData` interface as the native
backends. This is the closest equivalent to the old `anndata` R package: you can
call Python tools (e.g. `scanpy`) on the object while also using all `anndataR`
slot accessors from R.

```{r reticulate}
library(reticulate)
library(anndataR)

# Make scanpy available in the same Python environment that reticulate uses
py_require("scanpy")
sc <- import("scanpy")

# Read data with scanpy
# Note: anndataR will automatically wrap the resulting Python object in a ReticulateAnnData
url <- "https://cf.10xgenomics.com/samples/cell-exp/6.0.0/SC3_v3_NextGem_DI_CellPlex_CSP_DTC_Sorted_30K_Squamous_Cell_Carcinoma/SC3_v3_NextGem_DI_CellPlex_CSP_DTC_Sorted_30K_Squamous_Cell_Carcinoma_count_sample_feature_bc_matrix.h5"
ad <- sc$read_10x_h5("dataset.h5", backup_url = url)

# Use anndataR slot accessors from R
dim(ad)
head(ad$obs)
rowMeans(ad$X[1:10, ])

# Pass back to scanpy for preprocessing
# Note: anndataR automatically unwraps the Python object when calling scanpy functions.
sc$pp$filter_cells(ad, min_genes = 200L)
sc$pp$filter_genes(ad, min_cells = 3L)
sc$pp$normalize_per_cell(ad)
sc$pp$log1p(ad)

# Write from R when done
write_h5ad(ad, "path/to/output.h5ad")

# Migrate to a different anndataR backend
ad_mem <- ad$as_InMemoryAnnData()
# ad_hdf5 <- ad$as_HDF5AnnData(path = "path/to/disk_backed.h5ad")
# sce <- ad$as_SingleCellExperiment()
# seu <- ad$as_Seurat()
```

## Getting help with anndataR

- Documentation: <https://anndataR.scverse.org>
- Vignettes: `browseVignettes("anndataR")`
- Issue tracker: <https://github.com/scverse/anndataR/issues>
